\documentclass{article}

% Document Info
\author{Oliver Brotchie}
\title{A Catechumen's Handbook.}

\linespread{1.2}
\setlength{\parindent}{0pt}%
%\setlength{\parskip}{\baselineskip}%


% Packages
\usepackage[top=3cm,left=3cm,right=3cm,bottom=3cm]{geometry}
\usepackage{multicol}
\usepackage{ifthen}
\usepackage{xifthen}
\usepackage{xcolor}

\begin{document}
% Commands for service text
\newenvironment{service}{%
	\begin{trivlist}%
		\setlength\itemsep{1em}%
		}{%
	\end{trivlist}%
}

\newcommand{\alt}[1]{\begin{color}{red}#1\end{color}}

\newcommand{\altit}[1]{\alt{\textit{#1}}}

\newcommand{\breakcol}{%
	\vfill\null%
	\columnbreak%
}

\newcommand{\entry}[2][]{%
	\item\ifthenelse{\isempty{#1}}%
	{#2}%
	{\textbf{#1:} #2}%
}

\newcommand{\abr}[1]{%
	\begin{center}\vspace{-1.5em}\item\textit{#1}\end{center}%
}

\newcommand{\rest}{%
	\begin{center}\vspace{-2em}\item\alt{\textbf{. . .}}\end{center}%
}

\newcommand{\andagain}{\vspace{-0.1em}\abr{\alt{And again.}}}

\newcommand{\andsoafter}{ \alt{\textit{And so after each petition.}}}

\newcommand{\instruction}[1]{\item\alt{\vspace{1em}\textit{#1}\vspace{1em}}}

\newcommand{\psalm}[2]{%
	\item\textbf{Psalm #1}\\%
	#2%
}

\newcommand{\model}[2]{%
	\item\textbf{#1}%
	\item#2%
}

\newcommand{\cannon}[3][]{%
	\ifthenelse{\isempty{#1}}%
	{%
		\item\textbf{#2}
	}%
	{%
		\item\textbf{#2}
		\item\textit{#1}\vspace{-0.5em}%
	}%
	\item #3
}

\newcommand{\subheading}[1]{%
	\item \begin{center}{\textbf{#1}}\end{center}
}

% Common Prayers
\newcommand{\glory}{Glory to the Father, and to the Son, and to the Holy Spirit}
\newcommand{\bothnow}{Both now and forever, and to the ages of ages}
\newcommand{\glorybothnow}{\glory; \bothnow. Amen.}
\newcommand{\mercy}[1]{Lord, have mercy. \alt{(x#1)}}
\newcommand{\alleluia}{Alleluia, Alleluia, Alleluia. Glory to you, O God. \alt{(x3)}}
\newcommand{\ourfather}{%
	Our Father, in heaven, may your name be hallowed; your kingdom come;
	your will be done on earth as in heaven. Give us today our daily bread,
	and forgive us our debts, as we forgive our debtors. And do not lead
	us into temptation, but deliver us from the evil one.
}
\newcommand{\trisagion}{
	Holy God, Holy Strong, Holy Immortal, have mercy on us. \alt{(x3)}
	\item \glorybothnow
	\item All-holy Trinity, have mercy on us. Lord, cleanse us from our sins.
	Master, pardon our iniquities. Holy One, visit and heal our infirmities
	for your name's sake.
	\item \mercy{3}
	\item \glorybothnow
	\item \ourfather
}
\newcommand{\shortlitany}{
	\subheading{Short Litany}
	\entry[Priest]{
		Again and again in peace, let us pray to the Lord.
	}
	\entry[People]{
		Lord, have mercy.
	}
	\entry[Priest]{
		Help us, save us, have mercy on us, and keep us, O God, by your grace.
	}
	\entry[People]{
		Lord, have mercy.
	}
	\entry[Priest]{
		Commemorating our all-holy, pure, most blessed and glorious Lady,
		Mother of God and Ever-Virgin Mary, with all the Saints, let us
		entrust ourselves and one another and our whole life to Christ our God.
	}
	\entry[People]{
		To you, O Lord.
	}
	\entry[Priest]{
		For you are our God, and to you we give glory, to the Father, the Son
		and the Holy Spirit, now and for ever, and to the ages of ages.
	}
	\entry[People]{
		Amen.
	}
}

% Psalms

\maketitle
\pagebreak
\tableofcontents
\pagebreak
\section{The Service of Matins}
\subsection{Introduction}
\label{matins-intro}
\begin{service}
	\instruction{
		After the Dismissal of the Midnight Office the Priest,
		having made the customary metania to the Superior, or to his stall,
		enters the sanctuary and makes three bows before the holy Table.
		He puts on the Epitrachelion and stands in front of the Holy Table.
	}
	\begin{multicols}{2}
		\entry[Priest]{
			Blessed is our God, always, now and for ever, and to the ages of ages.
		}
		\entry[People]{
			Amen.
		}
		\instruction{
			The Priest blesses the incense and begins to cense the sanctuary and
			the whole church as usual.
		}
		\entry[Reader]{
			Glory to You, our God, Glory to You.
			\item Heavenly King, Paraclete, Spirit of truth, present everywhere
			filling all things, treasury of blessings and giver of life, come
			dwell in us, cleanse us of every stain, and, O Good One save our
			souls.
			\item \trisagion
		}
		\entry[Priest]{
			For yours is the kingdom, the power and the glory, of the Father,
			the Son and the Holy Spirit, now and for ever, and to the ages of ages.
		}
		\entry[Reader]{
			\mercy{12}
			\item \glorybothnow
			\item Come, let us worship and fall down before the King our God.
			Come, let us worship and fall down before Christ the King our God.
			Come, let us worship and fall down before Christ himself, the King, our God.
		}
	\end{multicols}
	\pagebreak
	\subsection{The Royal Psalms}
	\label{matins-royal}
	\instruction{
		The Royal Psalms are read before serving a Liturgy; for daily Matins, continue to Section
		\ref{matins-litany-before-six-psalms}.
	}
	\begin{multicols}{2}
		\psalm{19}{
			May the Lord hear you in the day of trouble; may the name of the God of Jacob
			shield you. May he send you his help from the holy place, and support
			you from Sion. May he remember your every sacrifice, and accept with favour
			your whole burnt offering. May the Lord give you your heart's desire, and
			fulfill your every purpose. We shall rejoice in your salvation, and be
			magnified in the name of our the Lord our God. May the Lord fulfill all
			your petitions. Now I know that the Lord has saved his Christ.
			He will hear him from his holy heaven; in mighty acts is the salvation
			of his right hand. Some put their trust in chariots and some in horses,
			but we will call on the name of the Lord our God. They were fettered
			and fell, but we have risen and been set upright. Lord, save the king,
			and hear us on the day we call upon you.
		}
		\psalm{20}{
			The king will rejoice in your power, O Lord, he will exult exceedingly
			in your salvation. You gave him his heart's desire; you did not deny
			him the request of his lips. For you came to meet him with blessings
			of goodness; you placed a crown of precious stones upon his head.
			He asked you for life, and you gave him length of days for age on age.
			Great is his glory because of your salvation; you will place on him
			glory and majesty. For you will give him blessing for age on age,
			and make him glad with the joy of your countenance. For the king
			puts his hope in the Lord, and through the mercy of the Most High
			he will not be shaken. May your hand light upon all your enemies,
			and your right hand find out all who hate you. You will make them
			like a blazing oven at the time of your presence. The Lord will
			confound them in his wrath and fire will devour them. You will
			destroy their offspring from the earth, and their seed from among
			the children of mankind. Because they intended evils against you,
			and devised plans by which they can in no way succeed. For you will
			put them to flight; among your remnants you will prepare their
			presence. Be exalted, Lord, in your power; we will sing and praise
			your mighty acts.
		}
		\entry[Reader]{
			\trisagion
		}
		\entry[Priest]{
			For yours is the kingdom, the power and the glory, of the Father,
			the Son and the Holy Spirit, now and for ever, and to the ages of ages.
		}
		\entry[People]{
			Amen.
		}
		\pagebreak
	\end{multicols}
	\pagebreak
	\subsection{Prayers and Litany before the Six Psalms}
	\label{matins-litany-before-six-psalms}
	\instruction{
		The priest must take care that he is standing beneath the Polyeleos
		to say this conclusion. He then enters the sanctuary, puts away the
		censer and takes his stand in front of the Holy Table.
	}
	\begin{multicols}{2}
		\entry[Reader]{
			Lord, save your people, and bless your inheritance, granting to faithful
			Christians victory over their enemies, and guarding your commonwealth
			by your Cross.
			\abr{Glory.}
			\item Lifted up on the Cross of your own will, to the new commonwealth
			that bears your name grant your mercies, Christ God; make your faithful
			people glad by your power, granting them victories over their enemies;
			may they have your help in battle: a weapon of peace, an invincible trophy.
			\abr{Both now.}
			\item Dread Champion who cannot be put to shame, do not despise our
			petitions, O Good One. All-praised Mother of God establish the commonwealth
			of the Orthodox, save your people and give them victory from heaven, for
			you gave birth to God, O only blessed one.
		}
		\rest
		\subheading{Litany before the Six Psalms}
		\entry[Priest]{
			Have mercy on us, O God, according to your great mercy. We pray you,
			hear and have mercy.
		}
		\columnbreak
		\entry[People]{
			\mercy{3} \andsoafter
		}
		\entry[Priest]{
			For our Archbishop \alt{N.}, for the honoured order of presbyters, for the diaconate
			in Christ, for all the clergy and the people, let us pray to the Lord.
			\item For our sovereign \alt{(lady/lord)}, \alt{N.}, the Royal Family, \alt{(his/her)} Government,
			and all in authority, let us pray to the Lord.
			\item For you, O God, are merciful, and love mankind, and to you we give
			glory, to the Father, the Son and the Holy Spirit, now and for ever and
			to the ages of ages.
		}
		\entry[People]{
			Amen.
		}
		\rest
		\entry[Reader]{
			In the name of the Lord, bless Father.
		}
		\entry[Priest]{
			Glory to the holy, consubstantial, life-giving and undivided Trinity,
			always, now and for ever, and to the ages of ages.
		}
		\entry[People]{
			Amen.
		}
	\end{multicols}
	\pagebreak
	\subsection{The Six Psalms}
	\label{matins-six-psalms}
	\instruction{
		And we begin the Six Psalms, listening with complete silence and compunction.
		During the Six Psalms the Priest reads the Twelve Morning Prayers.
		The Superior, or the designated reader, with devotion and fear of God reads.
		\footnote{
			The Six Psalms should be read by one reader, not divided between several.
			They are one of the gerontika, that is the parts normally read by the Superior,
			and should be allocated to a senior person, such a visiting priest. All should
			remain standing during the Six Psalms.
		}
	}
	\begin{multicols}{2}
		\entry[Reader]{
			Glory to God in the highest, and peace on earth, goodwill among men.\alt{(x3)}
			\item Lord, you will open my lips, and my mouth will declare your praise. \alt{(x2)}
		}
		\psalm{3}{
			Lord, why have those who afflict me been multiplied? Many rise against me.
			Many say to my soul, "There is no salvation for him in his God". But you,
			Lord, are my helper, my glory, and the one who lifts up my head. With my
			voice, I cried to the Lord, and he heard me from his holy mountain. I lay
			down and slept; I awoke, because the Lord will assist me. I will not be afraid
			of tens of thousands of people who surround me. Rise, Lord, save me, my
			God; because you have struck all who are vainly my foes. You have smashed
			the teeth of sinners. Salvation is the Lord's, and your blessing is upon your
			people.
			\andagain
			\item I lay down and slept; I awoke because the Lord will assist me.
		}
		\psalm{37}{
			Lord, do not rebuke me in your anger, nor chastise me in your wrath. For
			your arrows have been stuck fast in me, and you have laid a heavy hand upon
			me. There is no healing in my flesh on account of your wrath, nor peace in my
			bones on account of my sins. My iniquities flooded over my head; like a heavy
			burden they weighed heavily upon me. My wounds stank and festered on account of
			my foolishness. I was wretched and utterly bowed down; I went about mourning
			all day long. For my loins were filled with mockings, and there is no healing
			in my flesh. I was maltreated and humiliated exceedingly; I howled from the
			groaning of my heart. Lord, all my desire is before you, and my groaning was
			not hidden from you. My heart was troubled, my strength failed me; and even the
			light of my eyes had gone from me. My friends and my neighbours drew near and
			stood against me, and those nearest me stood far away. Those who sought my life
			used force, and those who sought evils for me spoke vanities, and meditated
			trickeries all day long. But I, like a deaf man, heard nothing; like one who is
			dumb, who does not open his mouth. So I became like one who cannot hear, and in
			whose mouth there are no rebukes. For in you, Lord, I have hoped; you will
			answer me, O Lord my God. For I  said: Let my foes never exult over me.
			When my foot slipped they crowed over me. For I am ready for blows and my pain
			is with me continually. For I will declare my iniquity and be troubled at my sin.
			But my enemies live and have prevailed over me, and those who hate me unjustly
			have been multiplied. Those who repay me with evils in place of good have
			slandered me, since I pursued goodness. Do not forsake me, Lord; my God, do not
			go far from me. Hasten to help me, Lord of my salvation.
			\andagain
			\item Do not forsake me, Lord; my God, do not go far from me. Hasten to help me,
			Lord of my salvation.
		}
		\psalm{62}{
			O God, my God, I watch for you at dawn; my soul has thirsted for you, my flesh
			how often in a desert, untrodden and waterless land. Thus I appeared before you
			in the holy place, to see your power and your glory. For your mercy is better than
			life; my lips will praise you. So I will bless you as long as I live, and in your
			name I will lift up my hands. May my soul be filled as with marrow and fatness;
			my mouth will praise you with joyful lips. When I remembered you upon my bed, every
			dawn I meditated upon you. For you became my helper, and in the shelter of your
			wings I shall rejoice. My soul clung to you; and your right hand upheld me.
			But those who vainly sought my life will go down to the deepest parts of the earth,
			they will be delivered to the power of the sword, they will be portions for foxes.
			But the king will rejoice in God, and all who swear by him will be praised; for
			the mouths of those who speak iniquities have been stopped.
			\vspace{-1em}
			\andagain
			\item Every dawn I meditated upon you, for you became my helper, and in the shelter
			of your wings I shall rejoice. My soul clung to you; and your right hand upheld me.
		}
		\rest
		\entry{
			\glorybothnow
			\item \alleluia
			\item \mercy{3}
			\item \glorybothnow
		}
		\rest
		\psalm{87}{
			Lord, God of my salvation, I have cried by day, and by night also before you.
			Let my prayer come into your presence, and turn you ear to my supplication.
			For my soul was filled with evils, and my life drew near to Hell. I was reckoned
			among those that go down to the Pit; I became as one that has no help, free among
			the dead. Like the slain that sleep in the tomb, whom you remembered no more; and they
			were cut off from your power. They placed me in the lowest Pit; in darkness and in the
			shadow of death. Your wrath lay heavy upon me, and you brought all your waves against
			me. You put my friends far from me, and made me an abomination to them. I was
			handed over and I came out; my eyes grew weak because of my beggary. I cried to you,
			Lord, all the day; I stretched out my hands towards you. Will you work wonders for
			the dead; or will physicians raise them up again to praise you? Will any declare your
			mercy in the tomb, or your truth in the place of destruction? Will your wonders be
			known in the dark, or your righteousness in a land that has been forgotten? But to you,
			Lord, I have cried, and in the morning my prayer will come before you. Why have you
			rejected my soul, Lord; why do you turn away your face from me? I am poor and in toils
			since my youth; exalted, I was humbled and brought to despair. Your fierce anger
			overwhelmed me, and your terrors appalled me. They surrounded me like water all day
			long; they closed in upon me from every side. Friend and neighbour you have put far
			from me, and my companions because of my misery.
			\andagain
			\item Lord, God of my salvation, I have cried by day; and by night also before you.
			Let my prayer come into your presence, and turn your ear to my supplication.
			\footnote{
				The priest comes from the Altar and reads the morning prayers quietly.
			}
		}
		\psalm{102}{
			Bless the Lord, O my soul, and all that is within me, bless his holy name.
			Bless the Lord, O my soul, and forget none of his benefits. He forgives all your
			iniquities; he heals all your diseases; he redeems your life from corruption
			and crowns you with mercy and compassion; he satisfies your desire with
			good things; your youth will be renewed like an eagle's. The Lord performs
			deeds of mercy and judgment for all who are oppressed. He made known his
			ways to Moses, his wishes to the children of Israel. The Lord is compassionate
			and merciful, long-suffering and full of mercy. He will not always be wrathful,
			nor will he be enraged for ever. He has not dealt with us according to our
			iniquities, nor rewarded us according to our sins. For, like the height of
			heaven above the earth, the Lord has made his mercy prevail over those that
			fear him. As far as the east is from the west, so far has he taken our sins from
			us. As a father takes pity on his children the Lord has taken pity on those who
			fear him. For he knows of what we are made; has remembered that we are but dust.
			As for man, his days are but as grass, he will blossom like a flower of the
			field; for when a wind has passed over it, it will be no more and will know its
			place no longer. But the mercy of the Lord is for ever and ever upon those
			who fear him; and his justice upon their children's children, upon those who
			keep his covenant and remember his commandments to do them. The Lord
			has prepared his throne in heaven and his kingdom rules over all. Bless the
			Lord, all you his Angels, powerful in strength, who perform his word, and
			listen to the voice of his words. Bless the Lord, all you his Powers, his
			Ministers who do his will. Bless the Lord, all you his works. In every place
			of his dominion, bless the Lord, O my soul!
			\andagain
			\item In every place of his dominion, bless the Lord, O my soul!
		}
		\psalm{142}{
			Lord, hear my prayer; in your truth give ear to my supplication, and in your
			righteousness hear me. Do not enter into judgement with your servant, for in
			your sight no one living can be justified. For the enemy pursued my soul;
			humbled my life to the ground; made me dwell in darkness, like those for
			ever dead. My spirit in me grew weary, and my heart was troubled within
			me. I remembered days of old; I meditated on all your works. I made the
			works of your hands my meditation. I stretched out my hands towards you;
			my soul thirsted for you like a waterless land. Hear me swiftly, O Lord; my
			spirit has failed. Do not turn your face from me, or I shall be like those
			who go down to the Pit. Make me hear of your mercy in the morning, for I have
			hoped in you. Make known to me the way in which I should walk, for to you I
			have lifted up my soul. Deliver me from my enemies, Lord; I have run to you
			for shelter. Teach me to do your will, for you are my God. Your good Spirit
			will guide me in an upright land. For your name's sake, O Lord, you will give
			me life. In your justice you will bring my soul out of trouble; in your mercy
			slay my enemies, and destroy all those who afflict my soul, for
			I am your servant.
			\andagain
			\item In your justice hear me, O Lord; and do not enter into judgment with
			your servant. \alt{(x2)}
			\item Your good Spirit will guide me in an upright land.
		}
		\rest
		\entry{
			\glorybothnow
			\item \alleluia
		}
	\end{multicols}
	\pagebreak
	\subsection{Litany of Peace}
	\label{matins-litany-of-peace}
	\instruction{
		The Priest, having kissed the icon of the Lord, stands in front of the Holy
		Doors.
	}
	\begin{multicols}{2}
		\entry[Priest]{
			In peace, let us pray to the Lord.
		}
		\entry[People]{
			Lord, have mercy.\andsoafter
		}
		\entry[Priest]{
			For the peace from on high and for the salvation of our souls, let us pray to
			the Lord.
			\item For the peace of the whole world, for the welfare of the holy Churches
			of God, and for the union of all, let us pray to the Lord.
			\item For this holy house, and for those who enter it with faith, reverence and
			the fear of God, let us pray to the Lord.
			\item For our Archbishop \alt{N.}, for the honoured order of presbyters, for the diaconate
			in Christ, for all the clergy and the people, let us pray to the Lord.
			\item For our sovereign \alt{(lady/lord)}, \alt{N.}, the Royal Family, \alt{(his/her)} Government,
			and all in authority, let us pray to the Lord.
			\item For our father and superior, Priest-monk \alt{N.}, and all our brotherhood in
			Christ.
			\item For this holy monastery, for monastery, city, town and village, and for the
			faithful who dwell in them, let us pray to the Lord.
			\item For favourable weather, an abundance of the fruits of the earth, and
			temperate seasons, let us pray to the Lord.
			\item For those who travel by land, air or water, for the sick, the suffering,
			for those in captivity, and for their safety and salvation, let us pray to the Lord.
			\item For our deliverance from all affliction, wrath, danger and constraint, let
			us pray to the Lord.
			\item Help us, save us, have mercy on us, and keep us, O God, by your grace.
			\item Commemorating our all-holy, pure, most blessed and glorious Lady, Mother
			of God and Ever-Virgin Mary, with all the Saints, let us entrust ourselves and
			one another and our whole life to Christ our God.
		}
		\entry[People]{
			To you, O Lord.
		}
		\entry[Priest]{
			For to you belong all glory, honour and worship, to the Father, the Son and the
			Holy Spirit, now and for ever, and to the ages of ages.
		}
		\entry[People]{
			Amen.
		}
	\end{multicols}
	\instruction{After the Litany, the following is sung, according to the tone of the
		Troparion of the day, by the two choirs alternately:
	}
	\entry{
		The Lord is God, and has appeared to us. Blessed is he who comes in the
		name of the Lord.
		\item \altit{[A] Verse 1:} Give thanks to the Lord, for he is good; his mercy endures
		for ever.
		\item \altit{[B] Verse 2:} All the nations surrounded me, but in the name of the Lord
		I drove them back.
		\item \altit{[A] Verse 3:} This is the Lord's doing, and it is marvelous in our eyes.
	}
	\pagebreak
	\subsection{Readings of the Kathismata}
	\label{matins-kathisma}
	\instruction{
		Then we say the Troparion of the current feast, or of the Saint(s) of the day, together
		with the Theotokion for the day of the week in the corresponding Tone. The first Kathisma
		is then read. During daily Matins, we read the N'th Psalm of the Kathisma, where
		N is the numbering of tone of the week. When outside of Great Lent, the ordering of
		the Kathismata is as follows:
	}
	\begin{center}
		\begin{tabular}{c c}
			\alt{Day} & \alt{Kathisma} \\ [0.5ex]
			\alt{Su}  & II, III        \\
			\alt{M}   & IV, V          \\
			\alt{Tu}  & VII, VIII      \\
			\alt{W}   & X, XI          \\
			\alt{Th}  & XIII, XIV      \\
			\alt{F}   & XIX, XX        \\
			\alt{Sa}  & XVI, XVII      \\
		\end{tabular}
	\end{center}
	\rest
	\begin{multicols}{2}
		\instruction{
			We begin the reading of each Kathismata with:
		}
		\entry[Reader]{
			\mercy{3}
			\item \glorybothnow
		}
		\instruction{
			And end each with:
		}
		\entry[Reader]{
			\glorybothnow
			\item \alleluia
		}
		\instruction{
			After reading the first and second Kathismata, sessional hymns from the Menaion
			or Ochtoechos are sung.
		}
		\model{Model Sessional Hymn One}{
			Having been conceived in iniquities, I, the prodigal dare not gaze upon
			the height of Heaven. But trusting in Thy love for man, I cry: O God, be
			gracious unto me, and save me.
			\item If the righteous scarcely be saved, where shall I, the sinner, appear?
			I have not borne the burden and heat of the day, number me together with
			those of the eleventh hour, O God, and save me.
			\abr{Glory. Both now. \alt{Theotokion}}
			\item Possessing the Theotokos Mary as the indestructable wall, O ye faithful,
			come, let us worship and fall down before her; for she hath boldness with
			Him that was born of her, both to intercede and to save our souls from wrath
			and death.
		}
		\model{Model Sessional Hymn Two}{
			Thy fatherly embrace open wide to my quickly, for like the prodigal I
			have wasted my whole life. O Saviour, have regard to the unfailing wealth of
			Thy compassions, and do not distain a heart made poor as a beggar. For, to
			The, O Lord, do I cry out in compunction: I have sinned against Thee; save me.
			\item Thy judgment seat is dreadful, and just is Thy judgement; my deeds fill
			me with fear; but O merciful Saviour, do Thou Thyself swiftly come to redeem
			me from punishment; save me from the portion of the goats, O good Master,
			and account me worthy yet to stand at Thy right hand, O holy and righteous
			Judge.
			\abr{Glory. Both now. \alt{Theotokion}}
			\item Cast off the sleep of sloth, O my soul, and be wakeful; light up the
			shining lamp of repentance in splendour; now go fourth rejoicing from the
			confusion of earthly life, so to meet the great immortal Bridegroom, while
			saying: Cast me not away, O Word, but freely receive me, through Thy holy
			Mother's prayers.
		}
		\instruction{
			After the second sessional hymn, the fiftieth psalm is then read during daily Matins.
		}
		\psalm{50}{
			Have mercy on me, O God, in accordance with your great mercy. According
			to the multitude of your compassion blot out my offence. Wash me thoroughly
			from my wickedness, and cleanse me from my sin. For I acknowledge
			my wickedness, and my sin is ever before me. Against you alone I have
			sinned and done what is evil in your sight, that you may be justified in your
			words and win when you are judged. For see, in wickedness I was conceived
			and in sin my mother bore me. For see, you have loved truth; you have shown
			me the hidden and secret things of your wisdom. You will sprinkle me with
			hyssop and I shall be cleansed. You will wash me and I shall be made whiter
			than snow. You will make me hear of joy and gladness; the bones which have
			been humbled will rejoice. Turn away your face from my sins and blot out all
			my iniquities. Create a clean heart in me, O God, and renew a right Spirit
			within me. Do not cast me out from your presence, and do not take your Holy
			Spirit from me. Give me back the joy of your salvation, and establish me with
			your sovereign Spirit. I will teach transgressors your ways, and sinners will
			turn to you again. O God, the God of my salvation, deliver me from bloodshed
			and my tongue will rejoice at your justice. Lord, you will open my lips,
			and my mouth will proclaim your praise. For if you had wanted a sacrifice, I
			would have given it. You will not take pleasure in burnt offerings. A sacrifice
			to God is a broken spirit; a broken and a humbled heart God will not despise.
			Do good to Sion, Lord, in your good pleasure; and let the walls of Jerusalem
			be rebuilt. Then you will be well pleased with a sacrifice of righteousness,
			oblation and whole burnt offerings. Then they will offer calves upon your altar.
		}
	\end{multicols}
	\pagebreak
	\subsection{Recitation of the Nine Odes}
	\label{matins-nine-odes}
	\instruction{
		We then begin to read the Nine Odes. An Hermos is sung after Odes III, VI,
		VIII, and IX. During daily Matins, one verse from either the Feast or from
		the  Ochtoechos is read, followed by three verses of the day. Short litanies
		are read after Odes III and VI.
	}
	\begin{multicols}{2}
		\cannon[Let us sing to the Lord, for he is greatly glorified.]
		{Model Ode I}{
			Let us sing to the Lord, for he is greatly glorified; horse and rider he has
			cast into the sea. Helper and protector he has become for my salvation; he
			is my God and I will glorify him; my father's God and I will exalt him.
			\item Then the leaders of Edom and the rulers of the Moabites hastened;
			trembling took hold of them; all the inhabitants of Canaan melted away.
			Let fear and trembling fall upon them; by the greatness of your arm let
			them be turned to stone.
			\abr{Glory.}
			\item Till your people pass over, O Lord; till your people, whom you have
			purchased, pass over. The sanctuary which you have made, Lord; which your
			hands have prepared.
			\abr{Both now.}
			\item The Lord reigns over the ages and for ever and ever; for Pharaoh's
			horse went with the chariots and horsemen into the sea. But the children
			of Israel walked on dry land in the midst of the sea.
		}
		\cannon[Holy are you, O Lord, and my spirit sings your praise.]
		{Model Ode III}{
			My heart is established in the Lord, my spirit is exalted in my God; my
			mouth derides my enemies, I rejoice in your salvation. There is none holy
			like the Lord, and none righteous like our God. Do not boast or talk so
			very proudly; let no arrogance come from your mouth. For the Lord is a
			God of knowledge; and God prepares his ways.
			\item He grants the prayers of him who prays; and blesses the years of
			the righteous. The Lord will weaken his adversary; the Lord is holy.
			\abr{Glory.}
			\item Let not the wise man boast of his wisdom, nor the powerful boast
			of his power, nor the wealthy boast of his wealth. But let him who
			boasts boast of this; of understanding and knowing the Lord; and of
			executing judgement and righteousness in the midst of the earth.
			\abr{Both now.}
			\item The Lord has gone up to the heavens and thundered; being
			righteous he will judge the ends of the earth.He will give strength
			to our kings; and exalt the horn of his anointed ones.
		}
	\end{multicols}
	\pagebreak
	\item
	\shortlitany
	\rest
	\vspace{0.5em}
	\begin{multicols}{2}
		\cannon[Glory to your power O Lord.]
		{Model Ode IV}{
			O Lord, I heard your report and was afraid; Lord I considered your works
			and was amazed. When my soul is troubled, in anger you will remember mercy.
			God will come from Teman; and the Holy One from a shady, wooded mountain.
			His glory covered the heavens; and the earth was full of his praise.
			\item For the fig-tree will not bear fruit, nor will there be produce on
			the vines. The labour of the olive will deceive, and the fields yield no
			food.
			\abr{Glory.}
			\item The sheep have failed from lack of fodder, and there will be no
			oxen at the mangers. Yet I will rejoice in the Lord; I will be joyful in
			God my Saviour.
			\abr{Both now.}
			\item The Lord is my strength; he will perfectly station my feet. And
			he puts me on high places; for me to conquer with his song.
		}
		\cannon[O Lord our God give us peace.]
		{Model Ode V}{
			From nightfall my spirit is awake for you, O God; for your commands are
			light upon the earth. Learn righteousness, inhabitants of the earth.
			Let the impious be taken away, that he may not see the glory of the Lord.
			O Lord your hand was lifted up and they knew it not; but once they know
			they will be ashamed. Jealousy will seize an untaught people; and now
			fire devours their adversaries.
			\item Bring evils upon them, O Lord, bring evils upon them; the glorious
			ones of the earth. O Lord in affliction we remembered you, with a little
			affliction you chastised us.
			\abr{Glory.}
			\item As the woman in labour draws near the time of her delivery and
			cries out in her labour; so have we been to your beloved. We shall not
			fall, but they will fall, the inhabitants of the earth.
			\abr{Both now.}
			\item The dead shall arise and those in the graves shall be raised; and
			those in the earth shall rejoice. For the dew which comes from you is
			healing for them; but the land of the impious shall perish.
		}
		\cannon[As you did the prophet Jonas, save us, O Lord.]
		{Model Ode VI}{
			I cried to the Lord; and he heard me. You cast me into the depths of the
			heart of the sea; and the floods surrounded me. All your billows and your
			waves passed over me. Shall I look again towards your holy temple?
			\item Water encompassed me, even to my soul; the lowest deep closed round
			me. My head went down to the clefts of the mountains; I descended into the
			earth, whose bars are eternal barriers.
			\abr{Glory.}
			\item Let my life come up from corruption to you, O Lord my God. And let
			my prayer come to you, to your holy temple.
			\abr{Both now.}
			\item Those who preserve vain and foolish things have forsaken their own
			Mercy. I will pay you. O Lord, whatever I have vowed, for my salvation.
		}
	\end{multicols}
	\rest
	\shortlitany
	\pagebreak
	\begin{multicols}{2}
		\instruction{
			The Kontakion and Okos are then read if available, followed by the
			Synaxarion of the day.
		}
		\cannon[Our God and the God of our fathers, blessed are you.]
		{Model Ode VII}{
			Blessed are you, O Lord, the God of our fathers; and praised and glorified
			is your name for ever more. For you are righteous in all that you have done
			for us. And all your works are true, and your ways are right, and all your
			judgements are true; and judgements of truth you have executed in all that
			you have brought upon us. And upon Jerusalem the city of our fathers.
			\item Blessed are you, O Lord, the God of our fathers; to be praised and
			exalted above all for ever. And blessed is your glorious, holy name; to be
			praised and exalted above all for ever.
			\abr{Glory.}
			\item Blessed are you in the temple of your glory; to be praised and
			exalted above all for ever. Blessed are you who behold the deeps and sit
			upon the Cherubim; to be praised and exalted above all for ever.
			\abr{Both now.}
			\item Blessed are you on the glorious throne of your kingdom; to be praised
			and exalted above all for ever. Blessed are you in the firmament of heaven;
			to be praised and exalted above all for ever.
		}
		\cannon[Praise the Lord his works and exalt him above all for ever.]
		{Model Ode VIII}{
			Bless the Lord all you works of the Lord, praise the Lord, and exalt him
			above all for ever. Bless the Lord Angels of the Lord, praise the Lord,
			and exalt him above all for ever. Bless the Lord all you waters above the
			heavens, all you powers of the Lord, praise the Lord, and exalt him above
			all for ever. Bless the Lord sun and moon, stars of heaven, praise the Lord,
			and exalt him above all for ever.
			\item Bless the Lord all you birds of the air, beasts and cattle, praise the
			Lord, and exalt him above all for ever. Bless the Lord you sons of men. Let
			Israel bless the Lord, praise the Lord, and exalt him above all for ever.
			\abr{We bless Father, Son and Holy Spirit, the Lord.}
			\item Bless the Lord priests of the Lord and servants of the Lord, praise
			the Lord, and exalt him above all for ever. Bless the Lord spirits and souls
			of the righteous, holy and humble of heart, praise the Lord, and exalt him
			above all for ever.
			\abr{Both now.}
			\item  Bless the Lord Ananias, Azarias and Misael, praise the Lord, and exalt
			him above all for ever. Bless the Lord Apostles, Prophets and Martyrs, praise
			the Lord, and exalt him above all for ever. We bless Father, Son and Holy
			Spirit, we praise the Lord, and exalt him above all for ever.
		}
	\end{multicols}
	\pagebreak
	\subheading{
		The Magnificat
	}
	\instruction{
		After 8'th Ode is finished the Priest or Deacon will say the following
		in a loud voice. He then censes the whole sanctuary and church as usual.
	}
	\entry[Priest]{
		The Mother of God and Mother of the Light in hymns let us honour and magnify.
	}
	\vspace{2em}
	\entry{
		\begin{center}\textbf{Chorus}\end{center}
		Greater in honour than the Cherubim, and beyond compare more glorious than the
		Seraphim, without corruption you gave birth to God the Word, truly the Mother
		of God, we magnify you.
	}
	\vspace{1em}
	\entry{
		\alt{[A]} My soul magnifies the Lord, and my spirit has rejoiced in God my Saviour.
		\item\alt{[B]} For he has regarded the lowliness of his handmaiden, for behold from
		henceforth all generations will call me blessed.
		\item\alt{[A]} For he that is mighty has done great things for me, and holy is his name,
		and his mercy is on them that fear him from generation to generation.
		\item\alt{[B]} He has shown strength with his arm, he has scattered the proud in the
		imagination of their hearts.
		\item\alt{[A]} He has put down the mighty from their thrones and exalted the humble;
		he has filled the hungry with good things and the rich he has sent away
		empty.
		\item\alt{[B]} He has helped Israel his servant in remembrance of his mercy, as he spoke
		to our fathers, Abraham and his seed for ever.
	}
	\rest
	\vspace{0.5em}
	\cannon[Zachary blesses the birth of his child.]
	{Model Ode IX}{
		Blessed be the God of Israel, for he has visited and redeemed his people.
		And has raised up a horn of salvation for us in the house of his servant
		David. As he spoke by the mouth of his holy prophets, who have been since
		the world began. Salvation from our enemies and from the hand of all who
		hate us. To perform the mercy promised to our fathers, and to remember
		his holy covenant.
		\item The oath which he swore to our father Abraham, to grant us that
		without fear we, being delivered from the fear of our enemies. Might
		serve him in holiness and righteousness before him all the days of
		our life.
		\pagebreak
		\abr{Glory.}
		\item And you, child, will be called the prophet of the Highest, for you
		will go before the Lord to prepare his ways. To give knowledge of
		salvation to his people by the forgiveness of their sins through the
		tender mercy of our God.
		\abr{Both now.}
		\item  By which the Dayspring from on high has visited us, to appear to
		those who sit in darkness and in the shadow of death. To guide our feet
		into the way of peace.
	}
	\rest
	\vspace{-1.5em}
	\instruction{
		After the Hermos of the 9th Ode, we make a full prostration and sing
		the following.
	}
	\entry{
		It is truly right to call you blessed, who gave birth to God, ever-blessed
		and most pure and the Mother of our God. Greater in honour than the Cherubim,
		and beyond compare more glorious than the Seraphim, without corruption you
		gave birth to God the Word; truly the Mother of God, we magnify you.
	}
	\shortlitany
	\pagebreak
	\subsection{Exapostilaria}
	\instruction{
		When none are appointed in the Menaion or Ochtoechos.
	}
	\begin{multicols}{2}
		\subheading{Monday and Tuesday}
		\entry{
			As God you adorned the heavens with stars, and through your Angels You
			lead the whole earth to the light, O Creator of all things save those who hymn
			You.
			\item Let us all hymn John the Forerunner and Baptist of the Saviour, Prophet
			among the Prophets, nurseling of the desert, offspring of Elisabeth.
			\item Sweetness of the Angels, joy of the afflicted, protectress of Christians,
			Virgin Mother of the Lord, help me and deliver me from eternal tortures.
		}
		\subheading{Wednesday and Friday}
		\entry{
			The Cross is the guardian of the whole world; the Cross is the beauty of the
			Church; the Cross is the strength of kings; the Cross is the support of the faithful;
			the Cross is the glory of Angels and the wound of demons.
			\item Standing by your Cross, she who bore you without seed lamenting cried out,
			Alas sweetest child, how have you set from my sight? How have you been
			numbered among the dead?
		}
		\subheading{Thursday}
		\entry{
			Having sped to every place beneath the sun, you proclaimed the holy incarnation
			of Christ from a Virgin, truly turning the nations from error, enlightening them
			and teaching all to honour the holy Trinity, O Apostles of the Saviour.
			\item Let us all praise the chief shepherd and high priest, Nicholas the leader
			of the people of Myra; for many men he saved who were to die unjustly; and he appeared
			to the King in a dream with Avlavios, abolishing the unjust decree.
			\item Mary, censer of purest gold, having become the vessel of the uncontainable
			Trinity; in whom the Father was well-pleased, the Son dwelt, and the all-holy
			Spirit, having overshadowed you, showed you, O Maiden, to be Mother of
			God.
		}
		\subheading{Saturday}
		\entry{
			As God you have power over the living and the dead, and through your Angels you
			lead the whole earth to the light, O Creator of all things save those who sing
			your praise.
			\item In you we boast, Mother of God, and you we have as advocate before God:
			stretch out your invincible hand and trounce our foes, send forth your help to
			your servants from the holy place.
		}
	\end{multicols}
	\vspace{1em}
	\subsection{Praises}
	\psalm{148}{
		\alt{[A]} Praise the Lord from the heavens; praise him in the highest.
		To you praise due, O God.\item
		\alt{[B]} Praise him, all his angels: Praise him, all his Powers. To you
		praise is due, O God.\item
		\alt{[A]} Praise the Lord from the heavens; praise him in the highest. To you
		praise is due, O God.\item
		\alt{[B]} Praise him, all his angels: Praise him, all his Powers. To you
		praise is due, O God.\item
		\alt{[A]} Praise him, sun and moon; praise him, all you stars and light.\item
		\alt{[B]} Praise him, you highest heavens and you waters that are above the
		heavens. Let them praise the name of the Lord.\item
		\alt{[A]} For he spoke and they came into being; he commanded and they were
		created.\item
		\alt{[B]} He established them for ever and ever; he made an ordinance, and
		it shall not pass away.\item
		\alt{[A]} Praise the Lord from the earth; praise him, you sea-monsters and
		all deeps;\item
		\alt{[B]} Fire and hail, snow and ice and storm-wind; things that do his
		word.\item
		\alt{[A]} Mountains and all hills, fruiting trees and all cedars;\item
		\alt{[B]} Beasts of the wild, and all cattle, creeping things and winged
		birds;\item
		\alt{[A]} Kings of the earth and all peoples, rulers and all judges of
		the earth;\item
		\alt{[B]} Young men and maidens: old men and youths together, let them
		praise the name of the Lord; for his name alone has been exalted.\item
		\alt{[A]} His praise is above earth and heaven, and he will exalt the
		horn of his people.\item
		\alt{[B]} A hymn for all his holy ones; for the children of Israel,
		a people that draws near him.\item
	}
	\psalm{149}{
		\alt{[A]} Sing to the Lord a new song, his praise in the Church of the holy
		ones.\item
		\alt{[B]} Let Israel rejoice in him that made him, let the children of Sion
		be joyful in their king.\item
		\alt{[A]} Let them praise his name in the dance; let them sing his praise
		with timbrel and with harp.\item
		\alt{[B]} For the Lord is well-pleased with his people; he will exalt the
		meek with salvation.\item
		\alt{[A]} His holy ones will exult in glory, and rejoice upon their beds.\item
		\alt{[B]} The high praises of God in their mouths, and two-edged swords in
		their hands,\item
		\alt{[A]} To exact vengeance among the nations, punishments among the
		peoples.\item
		\alt{[B]} To bind their kings in fetters; and their nobles in shackles of
		iron.\item
		\alt{[A]} To execute upon them the judgement that is decreed; such glory
		will be for all his holy ones.\item
	}
	\psalm{150}{
		\alt{[B]} Praise God in his saints; praise him in the firmament of his
		power.\item
		\alt{[A]} Praise him for his mighty acts; praise him according to the
		greatness of his majesty.\item
		\alt{[B]} Praise him with the sound of the trumpet; praise him with lute
		and harp.\item
		\alt{[A]} Praise him with timbrel and dance; praise him with strings and
		pipe.\item
		\alt{[B]} Praise him with tuneful cymbals; praise him with loud cymbals.
		Let everything that has breath praise the Lord.\item
	}
	\abr{
		\alt{[A]} \glory.\\
		\alt{[B]} \bothnow.
	}
	\begin{multicols}{2}
		\entry[Reader]{
			To you glory is due, Lord our God, and to you we give glory, Father, Son and
			Holy Spirit, now and for ever, and to the ages of ages. Amen.
			\item Glory to God in the highest, and on earth peace, goodwill among men. We
			praise you, we bless you, we worship you, we glorify you, we give you thanks
			for your great glory. Lord, King, God of heaven, Father almighty: Lord,
			only-begotten Son, Jesus Christ and Holy Spirit. Lord God, Lamb of God, Son of
			the Father, who take away the sin of the world, have mercy on us; you take
			away the sins of the world. Receive our prayer, you who sit on the right hand
			of the Father, and have mercy on us. For you alone are holy, you alone are
			Lord, Jesus Christ, to the glory of God the Father. Amen.
			\item Every day I will bless you, and praise your name for ever and to the ages of
			ages. Lord, you have been our refuge from generation to generation.\\I said,
			Lord, have mercy on me, heal my soul, for I have sinned against you. Lord, I
			have run to you for refuge: teach me to do your will, for you are my God. For
			with you is the source of life: and in your light we shall see light. Continue
			your mercy towards those who know you.
			\item Grant, Lord, this day to keep us without sin. Blessed are you, O Lord, the God
			of our fathers, and praised and glorified your name to the ages. Amen. May
			your mercy, O Lord, be upon us, as we have put our hope in you. Blessed are
			you, Lord, teach me your statutes. Blessed are you, Master, make me understand
			your statutes. Blessed are you, Holy One, enlighten me with your statutes.
			Lord, your mercy is for ever; do not scorn the work of your hands. To
			you praise is due, to you song is due, to you glory is due, Father, Son and
			Holy Spirit, now and for ever, and to the ages of ages. Amen.
		}
	\end{multicols}
	\rest
	\subheading{Litany of Fervent Intercession}
	\entry[Priest]{
		Let us complete our morning prayer to the Lord
	}
	\entry[People]{
		Lord, have mercy.\andsoafter
	}
	\entry[Priest]{
		Help us, save us, have mercy on us and keep us, O God, by your grace.
		\item That the whole day may be perfect, holy, peaceful and sinless, let us ask of the
		Lord.
	}
	\entry[People]{
		Grant this, O Lord.\andsoafter
	}
	\entry[Priest]{
		An angel of peace, a faithful guide, a guardian of our souls and bodies,
		let us ask of the Lord.
		\item Pardon and forgiveness of our sins and offences, let us ask of the Lord.
		\item Things good and profitable for our souls, and peace for the world, let us ask
		of the Lord.
		\item That we may live out the rest of our days in peace and repentance, let us ask
		of the Lord.
		\item A Christian end to our life, painless, unashamed and peaceful, and a good defence before the dread judgement seat of Christ, let us ask.
		\item Commemorating our all-holy, pure, most blessed and glorious Lady, Mother
		of God and Ever-Virgin Mary, with all the Saints, let us entrust ourselves and
		one another and our whole life to Christ our God.
	}
	\entry[People]{
		To you, O Lord.
	}
	\entry[Priest]{
		For you are a God of mercy, compassion and love towards mankind,
		and to you we give glory, to the Father, the Son and the Holy Spirit, now and
		for ever, and to the ages of ages.
	}
	\entry[People]{
		Amen.
	}
	\entry[Priest]{
		Peace to all.
	}
	\entry[People]{
		And to your spirit.
	}
	\entry[Priest]{
		Let us bow our heads to the Lord.
	}
	\entry[People]{
		To you, O Lord.
	}
	\subheading{Prayer at the Bowing of Heads}
	\entry[Priest]{
		Holy Lord, dwelling on high and beholding things below
		and, with your eye that observes all, keeping watch over the whole creation,
		to you we have bowed the neck of our soul and body, and we beseech you, O
		Holy of Holies: Stretch forth your invisible hand from your holy dwelling and
		bless us all. And, as you are good and love mankind, pardon us if we have
		sinned in anything, voluntarily or involuntarily, granting us your blessings
		both of this world and of the world above. \altit{(In a low voice)}
		\item For yours it is to have mercy and to save us, O our God, and to you
		we give glory, to the Father, the Son and the Holy Spirit, now and for ever,
		and to the ages of ages. \altit{(Aloud)}
	}
	\entry[People]{
		Amen.
	}
	\pagebreak
	\subsection{Aposticha and Apolytikion}
	\subheading{Monday to Friday.}
	\entry{
		\alt{Refrain I:} We have been filled in the morning with your mercy, Lord, we have
		rejoiced and been glad. In all our day let us be glad, in return for the days
		when you humbled us, for the years when we saw evils. And look on your
		servants and on the work of your hands and guide their children.
	}
	\entry{
		\alt{Refrain II:} And may the splendour of the Lord our God be upon us, and direct
		the work of our hands for us, and direct the work of our hands.
	}
	\subheading{Saturday.}
	\entry{
		\alt{Refrain I:} Blessed are those whom you have chosen and taken, O Lord; they will
		dwell in your courts.
		\item \alt{Refrain II:} Their souls will dwell among good things.
		\item \alt{Refrain III:} And their memorial is from age to age.
	}
	\instruction{
		We then read the Idiomel, if there is one.
	}
	\entry[Reader]{
		It is good to give thanks to the Lord, to sing praises to your name, O Most
		High. To declare your love in the morning and your truth every night.
		\item\trisagion
	}
	\entry[Priest]{
		For yours is the kingdom, the power and the glory, of the Father,
		the Son and the Holy Spirit, now and for ever, and to the ages of ages.
	}
	\instruction{
		Then Apolytikion of the day. Glory. Both now. And the appropriate Theotokion.
	}
	\pagebreak
	\subsection{Conclusion}
	\subheading{Litany by the Priest}
	\entry[Priest]{
		Have mercy on us, O God, according to your great mercy. We pray you,
		hear and have mercy.
	}
	\entry[People]{
		\mercy{3} \andsoafter
	}
	\entry[Priest]{
		Also we pray for our Archbishop \alt{N.}, for the honoured order of presbyters, for the diaconate
		in Christ, for all the clergy and the people, let us pray to the Lord.
		\item For our sovereign \alt{(lady/lord)}, \alt{N.}, the Royal Family, \alt{(his/her)} Government,
		and all in authority, let us pray to the Lord.
		\item For you, O God, are merciful, and love mankind, and to you we give
		glory, to the Father, the Son and the Holy Spirit, now and for ever and
		to the ages of ages.
		\item For our father and superior, Priest-monk \alt{N.}, and all our brotherhood in
		Christ.
		\item Also we pray for mercy, life, peace, health, salvation, visitation, pardon and
		forgiveness of sins for the servants of God, all devout and Orthodox Christians, the benefactors,
		helpers, worshipers and pilgrims in this holy monastery; \textit{[and for the servants of God \alt{N.}]}
		and all who have asked for our prayers, unworthy though we are.
		\item Also we pray for the blessed and ever-remembered founders of this holy
		monastery, and for all our departed brothers and sisters, Orthodox believers,
		who have gone to their rest before us and who here and in all the world lie
		asleep in the Lord; \textit{[and for the servants of God \alt{N.}]}
		and that they may be pardoned all their offences, both voluntary and involuntary.
		\item Help us, save us, have mercy on us, and keep us, O God, by your grace.
		\item Commemorating our all-holy, pure, most blessed and glorious Lady, Mother
		of God and Ever-Virgin Mary, with all the Saints, let us entrust ourselves and
		one another and our whole life to Christ our God.
	}
	\entry[People]{
		To you, O Lord.
	}
	\entry[Priest]{
		For to you belong all glory, honour and worship, to the Father, the Son and the
		Holy Spirit, now and for ever, and to the ages of ages.
	}
	\entry[People]{
		Amen.
	}
	\pagebreak
	\subheading{Dismissal}
	\entry[Priest]{
		Wisdom.
	}
	\entry[People]{
		Holy Father give the blessing.
	}
	\entry[Priest]{
		Blessed is He Who Is, Christ our true God, always, now and for ever,
		and to the ages of ages.
	}
	\entry[Reader]{
		Amen. May the Lord God strengthen the holy and pure faith of devout
		and orthodox Christians, with his holy Church and this sacred Monastery, to
		ages of ages.
	}
	\entry[Priest]{
		Most holy Mother of God, save us.
	}
	\entry[Reader]{
		Greater in honour than the Cherubim, and beyond compare more glorious
		than the Seraphim, without corruption you gave birth to God the Word;
		truly the Mother of God, we magnify you.
		\item \glorybothnow
		\item \mercy{3}
		\item Holy Father give the blessing.
	}
	\instruction{
		And the Priest, lowering the Phelonion, continues.
	}
	\entry[Priest]{
		May Christ our true God, at the prayers of his most pure and holy Mother; of the holy,
		glorious and all-praised Apostles; of Saint \alt{N. (patron of the church)},
		of Saint \alt{N.}, whose memory we celebrate today; of the holy and righteous
		forebears of God, Joachim and Anne. And of all the Saints, have mercy on us and
		save us, for he is good and loves mankind.
	}
	\entry[People]{
		Amen.
	}
	\entry[Priest]{
		Through the prayers of our holy Fathers, Lord Jesus Christ our God,
		have mercy on us.
	}
	\entry[People]{
		Amen.
	}
\end{service}
\end{document}
